---
title: "Homework7"
author: "Oohita Dasari"
output: rmarkdown::github_document
---

* I have taken this dataset from `dslabs` datasets link "https://rdrr.io/cran/dslabs/man/". From all the datasets in the link I choose the data of contagious diseases in US states from "https://rdrr.io/cran/dslabs/man/us_contagious_diseases.html"
* Firstly, I installed package of "dslabs" by using the command `install.packages("dslabs")` and intalled dplyr , ggplot2 libraries.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("dslabs")
library(dslabs)
library(dplyr)
library(ggplot2)
```


* In order to read the data I used `data(us_contagious_diseases)` and then view the data using `View(us_contagious_diseases)`. 


```{r}

data(us_contagious_diseases)

?us_contagious_diseases

View(us_contagious_diseases)

str(us_contagious_diseases)
```

* The dataset contains six attributes namely "disease","state", "year" , "weeks_reporting","count", "population". `Year` is a date attribute, `disease` and `state` belong to categorical attribute. `weeks_reporting`,`count` and `population` belong to quantative attribute. 

* The `disease` attribute contains details of 7 types of diseases that are spread throught the US states. `state` attribute provides detaild of the particular state the disease is reported. `year` attribute gives the details of the year in which the disease is reported. `weeks_reporting` has the information of number of weeks that the count of diseased people is reported in that state. `count` attribute shows the number of people affected with that patricular disease in that particular year. `population` has the details of the total population of that state during that particular year.


# Exploratory Data Analysis

**Question1  : What is the dataset saying?**


* Initially I wanted to visualize the data to understand the data and notice the trends. 
* I created a plot of the in order to look for the distribution of the count of people affected by the 7 types of diseases in each US state.
* I choosed a stacked bar chart for this visualization which groups the data of particular state and shows the variation with respect to diseases.


```{r}
gg<-ggplot(us_contagious_diseases, aes(state, count))
gg + geom_bar(position = "stack", stat = "identity",aes(fill = disease))+
  scale_fill_brewer(palette = "Set1")+ 
  ggtitle("Overall count of diseased people in each state of US,all years")+ 
  theme(axis.text.x = element_text(angle = 90))
```

* From the plot generated I can notice that approximately 80% of the states (42 states) has their count of people effected by the diseases less than `1000000(1 million)` people.
* Only 20% on the US states(10 states) have their count of people effected by the diseases more than 1 million.  From which we can clearly understand that the disease rate is high in that 10 states.

----------------------
  
**Question 2  :What is the trend of diseases through years?**
  
* Now I wanted to visualize the rate of people affected by each variety of disease. 
* I grouped the data with similar years and similar diseases. Then I calculated the ratio of sum of count to sum of population and multipied with 10000 in order to calculate the rate at which the people are effected by the disease.


```{r}  
  us_contagious_diseases %>%
  group_by(year, disease) %>%
  summarize(rate = sum(count)/sum(population)*10000) %>%
  ggplot(aes(year, rate, color = disease)) +
  geom_line()+ ggtitle("Yearly analysis of Rate of people affected by all diseases (considering NA values)")
```

* From the above graph it can be observed that there is no trend before the year 1960, this might be because there are many NA values or zeros for the data during those years. Due to which, the graph couldn't plot the values for the rate because of the missing data in the population and count attributes as we used them for caluclating the rate. 

  
 
**Question 3 : What if NA values are excluded?**
* From the previous visualization, I found that the graph is inappropriate due to the NA values. Now, I wanted to exclude the NA values in order to visualize the trends.
* I used "filter(!is.na(population))" to exck=lude the NA values in the data. 

```{r} 
us_contagious_diseases %>%
  filter(!is.na(population)) %>%
  group_by(year, disease) %>%
  summarize(rate = sum(count)/sum(population)*10000) %>%
  ggplot(aes(year, rate, color = disease)) +
  geom_line()+ ggtitle("Yearly analysis of Rate of people affected by all diseases (excluding NA values)")
```



* It can be observed that after removing the NA values, the trend is clearly visible for the years before 1960 as well.
* So from the plot, we can see that the rate of people affected by measels is very high followed by rate of people affected by polio when compared to allother diseases.
* It can aso be noticed that, the rate of people affected by the diseases are gradually reduced with years after 1960 and reduced to minimum from the year 1980 onwards.



# Detailed Questions generated from EDA

**Question 4: How are the trends of diseases in states where the overall population affected by diseases is greater than 17.5 millio?**

* From the first visualization I found that the highest count of diseased population above 17.5 million is found in four states (Arkansas, New mexico, Pennsylvania, tennessee)
  
* So I want to visualize the trends of how the rate of population affected by diseases varied through years, for which I filtered the data of states.
  
  ***Plot for Arkansas:***
  
  * I used filter to get a subset of data for Arkansas and grouped the data of similar years and diseases. Now, I plotted line chart to observe the trend of people affected by various diseases in that state.
  
```{r}
  us_contagious_diseases %>% filter(state=="Arkansas", !is.na(population)) %>% 
  group_by(year, disease) %>%
  summarize(rate = sum(count)/sum(population)*10000) %>%
  ggplot(aes(year, rate, color = disease)) + 
  ggtitle("Trend of diseases in Arkansas through years")+
  geom_line()
```

  * As it is observed that the rate of people effected by the diseases is decreasing to minimum after the year 1970, I'm concentrating on the trends between the years 1950 and 1970.
  * In arkansas, the highest reported disease is `measles` with nearly 70% of the population during the period of 1950 to 1955. Later the rate got decreased to 10% of the population which is a drastic change.
  * Smallpox is not reported ater the year 1953(approximately)
  * Polio had the highest rate of 10% during the year 1950 and then it is gradually decreased to minimum value from around 1956.
  * Diseases like Mumps, Rubella, Hepatitis A, Pertussis are seemed to be reported from the year 1968 onwards. However,they are reported with minimum count of people.


  
  ***Plot for New Mexico:***
  
  * I used filter to get a subset of data for New Mexico and grouped the data of similar years and diseases. Now, I plotted line chart to observe the trend of people affected by various diseases in that state.
  
```{r}
  us_contagious_diseases %>% filter(state=="New Mexico", !is.na(population)) %>% 
  group_by(year, disease) %>%
  summarize(rate = sum(count)/sum(population)*10000) %>%
  ggplot(aes(year, rate, color = disease)) + 
  ggtitle("Trend of diseases in New Mexico through years")+
  geom_line()
```

  * As it is observed that the rate of people effected by the diseases is decreasing to minimum after the year 1970.Therefore, I'm concentrating on the trends between the years 1950 and 1970.
  * In New Mexico, the highest reported disease is `measles` with nearly 89% of the population during the period of 1955 to 1960. Later the rate got decreased to less than 3% of the population by the year 1960. There is no count reported during the period 1962 to 1964 which is unusual.Again,the rate of people effected by measles is minimised after 1967. 
  * Smallpox is not reported ater the year 1953(approximately)
  * New Mexico has the highest rate of people effected with Polio as around 27.5% during the year 1938 which is greater than that reported in Arkansas and then it is gradually decreased to minimum value from around 1956.
  * Unlike Arkansas, Pertussis seems to be reported more in New mexico from 1968 to 1975. Diseases like Mumps, Rubella, Hepatitis A are seemed to be reported from the year 1968 onwards. However,they are reported with minimum count of people.
  

  
  ***Plot for Pennsylvania:***
  
  * I used filter to get a subset of data for Pennsylvania and grouped the data of similar years and diseases. Now, I plotted line chart to observe the trend of people affected by various diseases in that state.

```{r}
  us_contagious_diseases %>% filter(state=="Pennsylvania", !is.na(population)) %>% 
  group_by(year, disease) %>%
  summarize(rate = sum(count)/sum(population)*10000) %>%
  ggplot(aes(year, rate, color = disease)) + 
  ggtitle("Trend of diseases in Pennsylvania through years")+
  geom_line()
```

  * As it is observed that the rate of people effected by the diseases is decreasing to minimum after the year 1970.Therefore, I'm concentrating on the trends before the year 1970.
  * In Pennsylvania, the highest reported disease is `measles` with nearly 137% of the population during the period of 1935 to 1940. Later the rate got decreased to less than 25% of the population by the year 1960. Again,the rate of people effected by measles is minimised after 1967. 
  * Smallpox is not reported ater the year 1953(approximately) and even for the years for which it is reported it has the least count reported. 
  * Pennsylvania has the highest rate of people effected with Polio as around 23% during the year 1940 which is greater than that reported in Arkansas bus less than that of New Mexico and then it is gradually decreased to minimum value from around 1956.
  * Diseases like Mumps, Rubella, Hepatitis A, Pertussis are seemed to be reported from the year 1968 onwards. However,they are reported with minimum count of people.It seems that some diseases are not reported during few years.


 
  ***Plot for Tennessee:***
  
  * I used filter to get a subset of data for Tennessee and grouped the data of similar years and diseases. Now, I plotted line chart to observe the trend of people affected by various diseases in that state.

```{r}  
   us_contagious_diseases %>% filter(state=="Tennessee", !is.na(population)) %>% 
  group_by(year, disease) %>%
  summarize(rate = sum(count)/sum(population)*10000) %>%
  ggplot(aes(year, rate, color = disease)) + 
  ggtitle("Trend of diseases in Tennessee through years")+
  geom_line()
```


  * As it is observed that the rate of people effected by the diseases is decreasing to minimum after the year 1970.Therefore, I'm concentrating on the trends before the year 1970.
  * In Tennessee, the highest reported disease is `measles` with nearly 78% of the population during the period of 1955 to 1960. Later the rate got decreased to less than 35% of the population by the year 1966. Again,the rate of people effected by measles is minimised after 1968. 
  * Smallpox is not reported ater the year 1953(approximately) . 
  * The highest rate of people effected with Polio is around 10% during the year 1944  and then it is gradually decreased to minimum value from around 1956.
  * Pertussis seems to be reported in Tenessee at the rate od 16% which is more than what is reported in New Mexico. Diseases like Mumps, Rubella, Hepatitis A are seemed to be reported from the year 1968 onwards. However,they are reported with minimum count of people.It seems that some diseases are not reported during few years.

* From all the above plots it can be observed that Measles used to be a very common disease before 1960 in all the selected four states of US with highest count of people effected by it. However, it is also observed that they handled it effectively after 1960.


  
**Question 5 : Will weeks_reporting attribute effect the visualization?**
  
  * Now, As it is observed in the above graphs that there is some sudden drop in the trends and there is no trend during few years.
  * Therefore, I wanted to observe, If the weeks_reporting attribute that describes the number of weeks in a particular year that the count of people effeted by a particular disease is reported  has any influence of the trends.
  * I plotted the graphs for all the four states of US for which the highest rate of population effeted by diseases is observed.
  
  
  ***Plot for Arkansas:***
  
```{r}  
  us_contagious_diseases %>% filter(state=="Arkansas" & weeks_reporting >= 10, !is.na(population)) %>% 
  group_by(year, disease) %>%
  summarize(rate = sum(count)/sum(population)*10000) %>%
  ggplot(aes(year, rate, color = disease)) + 
  ggtitle("Trend of diseases in Arkansas through years (considering weeks reporting)") +
  geom_line()

```


  ***Plot for New Mexico:***

```{r}  
  us_contagious_diseases %>% filter(state=="New Mexico" & weeks_reporting >= 10, !is.na(population)) %>% 
  group_by(year, disease) %>%
  summarize(rate = sum(count)/sum(population)*10000) %>%
  ggplot(aes(year, rate, color = disease)) +
  ggtitle("Trend of diseases in New Mexico through years (considering weeks reporting)") +
  geom_line()
```


 
  ***Plot for Pennsylvania:***

```{r}  
  us_contagious_diseases %>% filter(state=="Pennsylvania" & weeks_reporting >= 10, !is.na(population)) %>% 
  group_by(year, disease) %>%
  summarize(rate = sum(count)/sum(population)*10000) %>%
  ggplot(aes(year, rate, color = disease)) + 
  ggtitle("Trend of diseases in Pennsylvania through years (considering weeks reporting)") +
  geom_line()
```
  
  


  ***Plot for Tennessee:***

```{r}  
  us_contagious_diseases %>% filter(state=="Tennessee" & weeks_reporting >= 10, !is.na(population)) %>% 
  group_by(year, disease) %>%
  summarize(rate = sum(count)/sum(population)*10000) %>%
  ggplot(aes(year, rate, color = disease)) + 
  ggtitle("Trend of diseases in Tennessee through years (considering weeks reporting)") +
  geom_line()
```

* From the above plots, it can observed that there is not much impact of weeks_reporting on the visualization. One observation can be noticed in the case of New Mexico data where we observed some missing data during the year 1960 to 1965. When weeks_reporting less than 10 in a year are excluded it can be observed that the plot has connectivity.